RED WINE ANALYSIS BY TEAGAN BLACK
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE,packages}
# Load all of the packages used for analysis
install.packages("ggplot2", dependencies = T, repos = "http://cran.us.r-project.org") 
install.packages("knitr", dependencies = T, repos = "http://cran.us.r-project.org")
install.packages('GGally', repos = "http://cran.us.r-project.org")

install.packages('MASS', repos = "http://cran.us.r-project.org")
install.packages('memisc', repos = "http://cran.us.r-project.org")
library(GGally)
library(MASS)
library(ggplot2)
library(memisc, lib.loc = "C:/Users/butte/Documents/R/win-library/3.4")
library(knitr)
```


```{r echo=FALSE, message=FALSE, warning=FALSE,Load_the_Data}
#create the database from the csv file and read out the column names


wine <- read.csv('wineQualityReds.csv')
names(wine)

```

> 

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots1}
summary(wine)

```

This shows the variations in the data. There seems to be some outliers -
for example in sulphates total sulfur dioxide. Let's look at the 
characterisitcs of a seemingly "normal" quality - alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots2}
ggplot(aes(x = alcohol), data =wine, binwidth = 0.2)+
  geom_histogram()

```

We can see that alcohol content seems positively skewed with the most samples between 9%
and 11% per volume. It would be interesting to see if the extreme cases (less than 9 or
more than 11) affect quality. Let's look at pH next.


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots3}
qplot(x = (pH), data =wine)

```

pH seems quite normally distributed, with most values falling between 3.0 and
3.5. There seem to be some "extremes", with some readings around 2.0 and some
above 4.0! Although this relates to the summary data, this is more acidic than I was 
expecting (without having done any research), given that 7.0 is neutral. My 
intuition is that very acidic wines (less than 3.0) would negatively affect quality, 
so it would be good to test this. 


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots4}
qplot(x = (volatile.acidity), data =wine)
qplot(x = (fixed.acidity), data =wine)
qplot(x = citric.acid, data =wine)

```

The three acidity values seem to have different characteristics. Volatile acidity
(measures acetic acid) seems to be mostly normally distributed with a slight tail
to the right whilst fixed acidity (measure tartartic acid) is more skewed to the
left. Citric acid (measures citric acid) is definitely skewed to the left, with 
most values being equal to 0. This would make a log transformation untenable. It
would be interesting to explore if there is any relationship between these three
characteristics. Let's see if the square root transform to the  volatile and fixed
acidity produces a more normal distribution.


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots5}
qplot(x = sqrt(volatile.acidity), data =wine)
qplot(x = sqrt(fixed.acidity), data =wine)

```

That seems to be more normal. Another three characteristics which could be related are the free
sulphites, total sulphites and sulphates.


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots6}
qplot(x = (free.sulfur.dioxide), data =wine, binwidth = 1)
qplot(x = total.sulfur.dioxide, data =wine, binwidth = 2)
qplot(x = sulphates, data =wine, binwidth = 0.025)

```

As I expected, the free sulfur dioxide and total sulfur dioxide data seems to have a 
very similar structure. The sulphates data is more normally distributed but has a tail 
to the right, with some extreme numbers above 1.5. Let's look at those extremes.


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots7}
subset(wine, wine$sulphates >1.5)

```

There seem to be 8 instances where the sulphatesis greater than 1.5. It seems
interesting that wines number 87 and 92 have exactly the same data but this could
just be a very similar wine. Wine number 93 has very similar, but not the same, data.
Apart from all the wines being average in quality. There does not seem to be other
similar characteristics. Let's look at the distribution of qualtiy to get a feel for
if having a subset of average wine is likely.


```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots8}
qplot(x = factor(quality), data =wine)


```

There does not seem to be any quality ratings below three or above eight. This could be
because the quality is the median of three scores, so this will inherently take away 
extreme cases. Given that it looks like the majority of scores are either 5 or 6, in the
above subset, it does not seem unlikely we could have picked 8 random scores of 5 or 6 
(although this would need to be statistically tested). The only other variables we have not
looked at are chlorides, density and sugar content

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots9}
#the lg10 of chlorides and residual sugar was taken to try coerce the data to be more normal
qplot(x = log10(chlorides), data =wine)
qplot(x = density, data =wine)
qplot(x = log10(residual.sugar), data =wine)

```

It is interestnig that the residual sugar levels and chloride levels have a similar
shape. Density is concentrated around 1.0, which makes sense as that is close to the
density of water. The documentation says the density is related to the residual sugar
content and alcohol content - it might be interesting to see if there is a relationship
there.

# Univariate Analysis

### What is the structure of your dataset?
This dataset explores 1,599 Portuguese Vinoh Verde red wine characteristics 
and their quality. "X" is the ID of the wine and "Quality" is a rating from 
0-10 (as assessed by the median of three wine experts), with 0 being bad and 
10 being incredible. The other data is all characteristics of the wine and 
are all measurable quantitative qualities (as opposed to qualitative, for 
example flavour. Most of the wines are of a quality between 4 and 6.
There seem to be some outliers in the sulpahtes, residual sugar and citric
acid data.

### What is/are the main feature(s) of interest in your dataset?

Whilst it would be ideal to be able to predict quality with some of the 
characteristics such as acidity or sulphate content, the limited range in the
quality of the wines might limit how statistically significant this is.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

What might be more interesting is to see how different characteristics affect
each other - for example, does density have anything to do with residual
sugar content? Or if the different types of acidity can affect each other.

### Did you create any new variables from existing variables in the dataset?

No

### Of the features you investigated, were there any unusual distributions? \

Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Yes, I took the log10 of the residual sugars and chlorides to get a less
skewed and more normal distribution.

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots}
#create a subset of data without the numbers of the wine as this is superfluous for ggpairs
wine_for_gg_pairs <- wine
wine_for_gg_pairs$X <- NULL
ggcorr(wine_for_gg_pairs, hjust = 0.5, size = 3)
```

Looking at all the pairs of plots (without number), most of the correlation numbers
seem fairly low. In fact, the only correlation above 0.5 is citric acid to fixed acidity
and total sulfuric acid to free sulfuric acid. Disappointingly, most of the qualities
do not seem to correlate with quality. Alcohol content has the highest correlation of
only 0.476. Most graphs seem patternless but there seems to be some sort of relationship
between density and fixed acidity, and between pH and fixed acidity, which I was not 
expecting. There is also potentially an interesting trend in the citric acidity and
density. I will take a look at these and perhaps some others to see if there are any 
patterns that are not apparent at first glance.

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots1}
ggplot(aes(x=density, y = fixed.acidity), data = wine)+
  geom_point(color='orange', 
             alpha=0.4)
```

Looking at fixed.acidity versus density, we can see there is what seems to be a linear
 relationship, although the data seems overfitted. Some alpha was added to try see 
 through this. There also seemed to be a relationship between citric acid and fixed acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots2}
ggplot(aes(x=(citric.acid), y = fixed.acidity), data = wine)+
  geom_point(color='blue', 
             alpha = 0.4)
```

This relationship seems less clear than the density one. There could perhaps be a non-
linear relationship. Let's see if we can use a transformation to get a more linear shape.


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots3}
ggplot(aes(x=(citric.acid), y = log10(fixed.acidity)), data = wine)+
  geom_point(color='blue', 
             alpha = 0.4)
```

Taking the log to base of 10 of the cube root of fixed acidity produces a more linear
shape. It is also noticeable that there are bunches of data points around whole numbers,
for example 0.00 and 0.50. This could suggest some inherent error in the data as I would
have expected the data to be more continuous (the bunches indicates discritization). THe
paper does not indicate how citric acid was physically measured but perhaps if the wine
was measured with different equipment that has different levels of precision. pH also seems
to affect fixed acidity so let's look at that next.


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots4}
ggplot(aes(x=(pH), y = fixed.acidity), data = wine)+
  geom_point(color='purple', 
             alpha = 0.4)
```

This also produces a non-linear shape so let's try a transform

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots5}
ggplot(aes(x=(pH), y = log10(fixed.acidity)), data = wine)+
  geom_point(color='purple', 
             alpha = 0.4)+
  geom_smooth(color = 'black')
```

Again, this is more linear. It would be interesting to see if we could build a model 
to predict fixed acidity. Adding in a smooth line to see the trend, it looks like our
data is fairly consistent in the middle but then gets distorted by extremes. It would 
be useful if we could predict quality so let's look at some of the relationships of the 
various qualities with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots6}
ggplot(aes( x = pH, y = quality), data = wine)+
  geom_point()
ggplot(aes( x = pH, y = quality), data = wine)+
  geom_jitter(alpha=0.4)
```

Looking at pH versus quality, it does not appear to be any obvious relationship. THe
data suffers from having most data point of quality 5 or 6. The first graph suffers from
over-fitting, so adding in some alpha and jitter gives a better graph. Looking at the few
data points for quality 3 and 8 shows that it could lower pH leads to a better quality.
Alcohol had some promising patterns in the ggpairs - let's look at that


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots7}

ggplot(aes( x = alcohol, y = quality), data = wine)+
  geom_jitter(alpha = 0.5, color='red')
```

Again using some jitter to try see through the discreteness of quality, it looks like 
there could be a weak relationship between alcohol and quality. I have heard wine drinkers
talk about the sulphate content in wine, so I wonder if that has any link to quality?


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots8}
ggplot(aes( x = quality, y = sulphates), data = wine)+
  geom_jitter(alpha = 0.5, color='dark green')
```

It does seem like there could be a relationship but the data is tightly bunched. Let's
exclude some of the extremes


```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots9}
ggplot(aes( x = quality, y = sulphates), data = subset(wine, wine$sulphates < 1.1))+
  geom_jitter(alpha = 0.5, color='dark green')
```

That is better - it definitely seems like there could be a vague linear relationship
here.


```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots3}
ggplot(aes( x = quality, y=alcohol), data = subset(wine, wine$alcohol <13.5))+
  geom_boxplot(aes(group=quality))+
  ggtitle("Alcohol content of Varying Quality Red Wines")+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))
```

It seems that a higher average alcohol content leads to better wine but it is noticeable
that this does not hold true for lower quality wines.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Unfortunately it is hard to see a significant relationship between quality
and other characterisitcs due to the limited range of samples beyond quality
of 4 and 6. Although you can see a potential relationship in the sulphates data, 
this really needs more data points to show a high correlation.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

What was unexpectedly interesting was the relationship between fixed acidity
and density, pH and citric acid. There seems an almost linear relationship
between the log of fixed acidity and those three characteristics

### What was the strongest relationship you found?

Between fixed acidity and citric acid (0.6 correlation coefficient)


# Multivariate Plots Section

It seemed like there were many relationships between fixed acidity and other
qualities so I am going to try build a clearer picture of fixed acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots}
ggplot(aes(x=(pH), y = log10(fixed.acidity), color = density), data = wine)+
  geom_point()+
  scale_color_continuous(low = '#4A235A',
                         high = '#D7BDE2')+
  ggtitle("pH, Fixed Acidity and Density of Red Wines")+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))


ggplot(aes(x=(pH), y = log10(fixed.acidity), color = citric.acid), data = wine)+
  geom_point()+
  scale_color_continuous(low = '#F6DDCC',
                         high = '#6E2C00')+
  ggtitle("pH, Fixed Acidity and Citrc Acidity of Red Wines")+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))+
  theme_dark()
```

Looking at fixed acidity, pH and citric acid, it seems the lower the citric
acid, the lower the log of fixed acidity but across all pHs. Density also
has a similar pattern. I wonder what this will look like cut by quality


```{r echo=FALSE, message=FALSE, warning=FALSE,Multivariate_Plots1}
ggplot(aes(x=(pH), y = log10(fixed.acidity), color = density), data = wine)+
  geom_point()+
  scale_color_continuous(low = '#4A235A',
                         high = '#D7BDE2')+
  facet_wrap(~quality)+
  ggtitle("pH, Fixed Acidity and Density of Red Wines over Different Qualities")+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))


ggplot(aes(x=(pH), y = log10(fixed.acidity), color = citric.acid), data = wine)+
  geom_point()+
  scale_color_continuous(low = '#F6DDCC',
                         high = '#6E2C00')+
  facet_wrap(~quality)+
  ggtitle("pH, Fixed Acidity and Citric Acid of Red Wines over Different Qualities")+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))+
  theme_dark()
```

THe first thing I notice is just how few points we have for the wines of
quality 3 or 8! But even between the other qualities, the graphs show a similar
patter ie. the distribution of pH, fixed acidity and density is similar for
different quality wines. This is similar for citric acid. Let's look at the
difference in alcohol content for different quality wines

 Whilst the relationship between quality
and the other characteristics seems weak, there is definitely some correlation between 
fixed acidity and pH, density and citrc acid so let's try build a model.

```{r echo=FALSE, message=FALSE, warning=FALSE,Multiivariate_Plots4}

m1 <- lm(I(log10(fixed.acidity)) ~ I(pH), wine)
m2 <- update(m1, ~ . + citric.acid)
m3 <- update(m2, ~ . + density)
mtable(m1,m2,m3)
```

Looking at the R squared value, these three variables account for
77% of the variability in fixed volatility. Trying to add other
variables does not significantly change this so I will leave the model
here.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I continues to look at the relationship between fixed acidity, density, pH
and citric acid. When combined in a linear model, density, pH and citric
acid can be used to account for roughly 77% of the variability in fixed
acidity (tartic acid). Of this citric acid comprised about 50%.

### Were there any interesting or surprising interactions between features?

The facet wrap graphs confirmed my suspicions that quality was quite 
unrelated to the changes in fixed acidity, citric acid and pH.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created a model. The model is strong in that it is based on a
large sample size but could only be used for Portguese wines from the 
Vinoh Verde grape and even then will still only predict with an R-
squared of 0.77, which is not very accurate.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_One}

ggplot(aes( x = quality, y = sulphates), data = subset(wine, wine$sulphates < 1.1))+
  geom_jitter(alpha = 0.5, color='dark green')+
  ggtitle("Sulphates content of Red Wines by Quality")+
  xlab('Quality')+
  ylab('Sulpahtes (potassium sulphate - g/dm3)')+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))

```

### Description One

This plot is interesting as it was one of the clearest indicators of quality, even
though it is not actually very clear. Because of the discreteness of the quality, jitter
was used to try "see" a more spread out pattern. It also demonstrates the extreme 
concentration of data points on wines of quality 5 and 6. It is also perhaps a reflection
of what I hoped to achieve when starting this - to find a model to predict quality. What
I ended up with was mostly plots like this but some clearer plots on fixed acidity!

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Two}

ggplot(aes(x=(pH), y = log10(fixed.acidity), color = citric.acid), data = wine)+
  geom_point()+
  scale_color_continuous(low = '#F6DDCC',
                         high = '#6E2C00',
                         name="Citric Acid (g/dm^3)\n")+
  ggtitle("pH, Fixed Acidity and Citric Acidity of Red Wines")+
  xlab('pH')+
  ylab('log10(Fixed Acidity - tartic acid (g/dm^3))')+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))
```

### Description Two

This graph shows a fairly clear relationship between pH and the
log of fixed acidity and was the basis for my decision to build a 
linear model. The color element also shows a relationship with
citric acid, which is interesting but that the more tartic acid 
(fixed acidity), the more citric acidity.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Three}

ggplot(aes(x=density), data = wine)+
  geom_histogram()+
  ggtitle("Density Distribution of Red Wines")+
  xlab('Density (g/cm^3)')+
  ylab('Frequency')+
  theme(plot.title = element_text(lineheight=.8, face="bold", hjust = 0.5))
```

### Description Three

This graphs shows a normal looking distribution of the densities of wine. It is 
interesting in that it is such a good example of a normal distribution! It is 
also interesting in that the median density is below that of water (1g/cm^3).

# Reflection

Going through the data analysis of the red wine samples, I expected to focus
mostly on factors that affected quality. Once I started diving into the data
it became apparent that there was not a wide enough variety in the quality
of wines to notice any patterns (or perhaps because of the calculation method for
determining the quality a wine received).

However this led to the surprising relationship between fixed volatility and
pH, density and citric acid content. It was really interesting to see the
graphs evolve from 1-D to 2-D, where you can see a pattern emerging, and finally
to add some color to the matter.

I would like to expand the dataset to include wines of different regions, as I
think this will allow for a greater range in all the characterisitcs, which
might lead to some insightful discoveries. The one thing that would also be
interesting to look at, is the relationship between pH and a combination of the
acidities recorded.
